Gaps in stellar streams as a result of globular cluster fly-bys

### 2025-05-23 paper/main.tex
removed name for highlighting green text

  \textbf{ \salvatore{\citet{2017MNRAS.470...60E} is another work that discusses the impact of globular clusters in producing gaps on the Pal~5 stream. In section~6.3 of their article, \citet{2017MNRAS.470...60E} discuss the fact that globular clusters with masses greater than $10^6~\textrm{M}_\odot$ are rare in the Galaxy, especially at distances compatible with the orbit of Pal~5, while -- based on previous works -- they estimate that sub-halos of dark matter of similar mass are at least three times greater in number. This difference leads them to conclude that the gaps they report in the Pal~5 streams are more likely to be induced by dark matter sub-halos than by globular clusters. 
  However, as we show in this paper, even clusters with masses below $10^6~\textrm{M}\odot$ can produce gaps. The clusters perturbing the Pal~5 stream are clusters with masses up to 10 times lower than those considered by \citet{2017MNRAS.470...60E}. It is thus possible that \citet{2017MNRAS.470...60E} have underestimated the impact of globular clusters' close passages on a stream such as Pal~5. We can derive from \citet{2016MNRAS.457.3817S}'s third figure that a more concentrated system delivers a higher velocity kick but over a shorter distance, while a more spread out system will affect more stars yet perturb them less. It is more difficult, however, to make a comparative analysis between the role of clusters and sub-halos at equal stellar mass. Being less dense than globular clusters, sub-halos should produce less intense perturbations on streams (for fixed impact parameters and relative velocities). A systematic comparative analysis to quantify the role of clusters and sub-halos in producing gaps and perturbations in stellar streams still needs to be done.}}

@@

    \salvatore{Fig.~\ref{fig:decoherence_shocked_outflow} dissects the more massive stream of Fig.~\ref{fig:gap_no_more} and instructs us why the gap does not persist. The bottom panel shows the total density distribution, and dissects it into two profiles: stars that either are \textit{leaked} or \textit{shocked} out of the cluster. By shocked, we mean the stars that are lost during pericenter passages. Tidal forces generally scale with $r^{-3}$, thus significantly strengthening through their pericenter passages. See Fig.~3 of \citet{2004AJ....127.2753D} for an illustration of the magnitude of the tidal forces along an orbit. A consequence of the pericenter passages is that many more stars escape, fleeing with a higher mean velocity and velocity dispersion, which can be visualized in the \textit{Christmas tree} illustration in Fig.~A.3 of \citet{2012A&A...546L...7M}. We consider the continous outflow of stars between pericenter passages to be \textit{leaked} stars. We identify pericenter passages by computing each local minimum of the cluster's galactic-centric 3D radius with \texttt{scipy.signal.find\_peaks}. The duration of each shock is: $t_{\textrm{shock}}$ = 2$\sqrt{r/|\nabla\Phi|}$, where $r$ is the galactocentric radius and $|\nabla\Phi|$ is the magnitude of the acceleration from the galactic potential. Stars within this range about a given pericenter passage are all given the same color and shown as individual groups in the first, second, third, and fourth panels of Fig.~\ref{fig:decoherence_shocked_outflow}.}

@@

    \salvatore{It is important to note that within a single packet, the stars self-segregate. The position of the density peak is at $\langle v \rangle t$, while the faster stars are ahead and the slower ones are behind. The packets are separated by a distance of $\langle  v\rangle T$, where $T$ is the orbital period. Regarding our simulations, since the mass and radius of the host are held constant, the mean velocity and velocity dispersion are the same across all packets. Secondly, the density of each packet decreases since the easier-to-remove stars are stripped early on. }

    \salvatore{Consider an impact that happens at $x\prime$. The distance from the impact site to each packet is: $\delta x= x\prime - \langle v\rangle \left(t-n T\right)$, where $n$ is the packet index, with $n=0$ being the first packet escaped at the first pericenter passage in the simulation. Note that $\delta x\propto n$, and velocity segregation means $v\propto x$. Therefore, the different parts of the separate packets are hit at a given impact site. The earlier packets are hit in regions where the stars have lower velocities than the later packets. As a result, the gaps in the later packets quickly overtake those from the earlier packets. As a result, the gaps across the packets go out of phase and interfere destructively.}

	  \salvatore{The aforementioned analysis highlights the complexity of the conditions on the stream that are necessary for allowing gap formation. Despite this, in the attempts to gain further preliminary insight, we can create an even simpler consideration of a gap created in two different pockets, considering the gaps' drift rates and width, ignoring their density contrasts. If we consider two packets with the same dimensions $\sigma_x$ and an offset of $\delta x$, the gaps are out of phase when the separation is greater than the width: $1\sim<\sigma_x/\delta x$. If $\delta x$ grows much faster than the gap can develop, then the gap is never visible. If the gaps grow in size much faster than their drift rate, then they persist and continue to grow in significance. In the simulations presented in this analysis, we only observe the two extremes and do not produce a case where a significant gap is created to blur out. However, this is not necessarily the general case.}

    \salvatore{Nonetheless, this is a first look at considering the distribution of stream stars originating from shocked outflow, and they allow or disallow gap formation. Fuller considerations would include mass loss from internal dynamics, such as three-body encounters, and a non-constant mass loss rate due to changes in the density distribution resulting from internal mass evolution.}

>>>>

  \citet{2017MNRAS.470...60E} is another work that discusses the impact of globular clusters in producing gaps on the Pal~5 stream. In section~6.3 of their article, \citet{2017MNRAS.470...60E} discuss the fact that globular clusters with masses greater than $10^6~\textrm{M}_\odot$ are rare in the Galaxy, especially at distances compatible with the orbit of Pal~5, while -- based on previous works -- they estimate that sub-halos of dark matter of similar mass are at least three times greater in number. This difference leads them to conclude that the gaps they report in the Pal~5 streams are more likely to be induced by dark matter sub-halos than by globular clusters.

  However, as we show in this paper, even clusters with masses below $10^6~\textrm{M}\odot$ can produce gaps. The clusters perturbing the Pal~5 stream are clusters with masses up to 10 times lower than those considered by \citet{2017MNRAS.470...60E}. It is thus possible that \citet{2017MNRAS.470...60E} have underestimated the impact of globular clusters' close passages on a stream such as Pal~5. We can derive from \citet{2016MNRAS.457.3817S}'s third figure that a more concentrated system delivers a higher velocity kick but over a shorter distance, while a more spread out system will affect more stars yet perturb them less. It is more difficult, however, to make a comparative analysis between the role of clusters and sub-halos at equal stellar mass. Being less dense than globular clusters, sub-halos should produce less intense perturbations on streams (for fixed impact parameters and relative velocities). A systematic comparative analysis to quantify the role of clusters and sub-halos in producing gaps and perturbations in stellar streams still needs to be done.

@@

    Fig.~\ref{fig:decoherence_shocked_outflow} dissects the more massive stream of Fig.~\ref{fig:gap_no_more} and instructs us why the gap does not persist. The bottom panel shows the total density distribution, and dissects it into two profiles: stars that either are \textit{leaked} or \textit{shocked} out of the cluster. By shocked, we mean the stars that are lost during pericenter passages. Tidal forces generally scale with $r^{-3}$, thus significantly strengthening through their pericenter passages. See Fig.~3 of \citet{2004AJ....127.2753D} for an illustration of the magnitude of the tidal forces along an orbit. A consequence of the pericenter passages is that many more stars escape, fleeing with a higher mean velocity and velocity dispersion, which can be visualized in the \textit{Christmas tree} illustration in Fig.~A.3 of \citet{2012A&A...546L...7M}. We consider the continous outflow of stars between pericenter passages to be \textit{leaked} stars. We identify pericenter passages by computing each local minimum of the cluster's galactic-centric 3D radius with \texttt{scipy.signal.find\_peaks}. The duration of each shock is: $t_{\textrm{shock}}$ = 2$\sqrt{r/|\nabla\Phi|}$, where $r$ is the galactocentric radius and $|\nabla\Phi|$ is the magnitude of the acceleration from the galactic potential. Stars within this range about a given pericenter passage are all given the same color and shown as individual groups in the first, second, third, and fourth panels of Fig.~\ref{fig:decoherence_shocked_outflow}.

@@

    It is important to note that within a single packet, the stars self-segregate. The position of the density peak is at $\langle v \rangle t$, while the faster stars are ahead and the slower ones are behind. The packets are separated by a distance of $\langle  v\rangle T$, where $T$ is the orbital period. Regarding our simulations, since the mass and radius of the host are held constant, the mean velocity and velocity dispersion are the same across all packets. Secondly, the density of each packet decreases since the easier-to-remove stars are stripped early on.

    Consider an impact that happens at $x\prime$. The distance from the impact site to each packet is: $\delta x= x\prime - \langle v\rangle \left(t-n T\right)$, where $n$ is the packet index, with $n=0$ being the first packet escaped at the first pericenter passage in the simulation. Note that $\delta x\propto n$, and velocity segregation means $v\propto x$. Therefore, the different parts of the separate packets are hit at a given impact site. The earlier packets are hit in regions where the stars have lower velocities than the later packets. As a result, the gaps in the later packets quickly overtake those from the earlier packets. As a result, the gaps across the packets go out of phase and interfere destructively.

	  The aforementioned analysis highlights the complexity of the conditions on the stream that are necessary for allowing gap formation. Despite this, in the attempts to gain further preliminary insight, we can create an even simpler consideration of a gap created in two different pockets, considering the gaps' drift rates and width, ignoring their density contrasts. If we consider two packets with the same dimensions $\sigma_x$ and an offset of $\delta x$, the gaps are out of phase when the separation is greater than the width: $1\sim<\sigma_x/\delta x$. If $\delta x$ grows much faster than the gap can develop, then the gap is never visible. If the gaps grow in size much faster than their drift rate, then they persist and continue to grow in significance. In the simulations presented in this analysis, we only observe the two extremes and do not produce a case where a significant gap is created to blur out. However, this is not necessarily the general case.

    Nonetheless, this is a first look at considering the distribution of stream stars originating from shocked outflow, and they allow or disallow gap formation. Fuller considerations would include mass loss from internal dynamics, such as three-body encounters, and a non-constant mass loss rate due to changes in the density distribution resulting from internal mass evolution.
